---
title: "Getting started"
description: "Introduction to talk: HuggingFace transformers in R."
author: ""
opengraph:
  image: 
    src: "http://r-text.org/articles/text_files/figure-html/unnamed-chunk-5-1.png"
  twitter:
    card: summary_large_image
    creator: "@oscarkjell"
output: github_document #rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{text}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<a rel="me" href="https://mastodon.online/@oscarkjell">Oscar Kjell</a>
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
evaluate = FALSE
```
The *talk*-package uses Hugging Face transformers language models, natural language processing and machine learning methods to analyse audio recordings of talk and numerical variables.

This Getting Started tutorial is going through some central *talk* functions.  

In short the workflow includes to first transform audio recordings into text (speech-to-text) as well as embeddings (which can be used in downstream tasks, for example using textTrain functions in the text-package.

talk uses a **single Python environment** (`talkrpp_condaenv`), installed once
with `talkrpp_install()`. It contains the full stack, so all functions —
`talkText()`, `talkEmbed()`, `talkTranscribeDiarise()` and `talkEmbedSegments()`
— run in the same environment.

The code chunks below are not run when the vignette is built (they download
models and need the Python environment). Set `evaluate = TRUE` to run them
interactively.

```{r setup, eval = evaluate, warning=FALSE, message=FALSE}
library(talk)
```

### Setting up the Python backend

#### talkrpp_install(): install the Python backend
Creates the `talkrpp_condaenv` conda environment and installs the full talk
stack (transcription, embeddings, diarisation and segment embeddings). Run it
**once**.

```{r install, eval = evaluate}
talkrpp_install(prompt = FALSE)
```

#### talkrpp_initialize(): activate the environment
Activates an installed environment for the **current R session**. Required
before `talkText()` / `talkEmbed()`. (`talkTranscribeDiarise()` and
`talkEmbedSegments()` activate their environment automatically in a subprocess,
so they do not need this.)

```{r initialize, eval = evaluate}
talkrpp_initialize(
  condaenv         = "talkrpp_condaenv",
  refresh_settings = TRUE,
  save_profile     = TRUE
)
```

#### find_talkrpp_env(): check whether an environment exists
Returns `TRUE`/`FALSE` — useful for scripts that should install only when needed.

```{r find-env, eval = evaluate}
if (!find_talkrpp_env()) {
  talkrpp_install()
}
```

#### list_talkrpp_envs(): list the available environments
Returns the names of the Python environments on your system (all conda
environments, plus any virtualenvs) — handy for seeing which name to pass as
`condaenv` (to `talkrpp_initialize()` / `talkTranscribeDiarise()`) or `envname`
(to `talkrpp_install()`).

```{r list-envs, eval = evaluate}
list_talkrpp_envs()
#> [1] "base" "talkrpp_condaenv" ...
```

### Transcription and embeddings

The example audio files ship with the package.

```{r example-data, eval = evaluate}
wav_short   <- system.file("extdata", "test_short.wav",   package = "talk")
wav_diarise <- system.file("extdata", "test_diarise.wav", package = "talk")
```

#### talkText(): transcribe audio to text
Transcribes an audio file to **text** (speech-to-text) using a Whisper model.

```{r talktext, eval = evaluate}
text <- talkText(
  talk_filepaths = wav_short,
  model          = "openai/whisper-small"
)
text
```

#### talkEmbed(): mapping audio recordings to numbers using a HuggingFace language model
The `talkEmbed()` function transforms an audio file into a **single embedding**
(one numeric vector per file), which can be used in downstream tasks — for
example with the `textTrain()` functions in the text-package.

```{r talkembed, eval = evaluate}
embeddings <- talkEmbed(
  talk_filepaths = wav_short,
  model          = "openai/whisper-small",
  use_decoder    = FALSE
)
embeddings
```

#### talkTranscribeDiarise(): transcribe and split by speaker
Transcribes **and** diarises a recording — producing a transcript split by
speaker and timestamped. Returns a list with `$transcript` (a data.frame),
`$output_files`, and `$status`. Requires the diarise environment.

```{r diarise, eval = evaluate}
diar <- talkTranscribeDiarise(
  audio        = wav_diarise,
  num_speakers = 2
)

diar$status
diar$transcript
```

#### talkEmbedSegments(): one embedding per diarised segment
Unlike `talkEmbed()` (one embedding per whole file), `talkEmbedSegments()`
returns **one embedding per diarised segment**. It takes the audio plus a
diarised transcript (e.g. `diar$transcript` above, or a CSV path). Two backends
are available: Whisper (`"encoder"` / `"decoder"` / `"both"`) and WhiSPA.

```{r embed-segments, eval = evaluate}
# Whisper encoder embeddings, one row per segment
emb <- talkEmbedSegments(
  audio      = wav_diarise,
  transcript = diar$transcript,   # accepts the diarised output directly, or a CSV path
  model      = "whisper",         # or "whispa"
  embeddings = "encoder"          # "encoder" / "decoder" / "both" (whisper only)
)
emb

# WhiSPA speech-psychological embeddings
emb_whispa <- talkEmbedSegments(
  audio      = wav_diarise,
  transcript = diar$transcript,
  model      = "whispa"
)
```

### Function summary

| Function | What it does |
|---|---|
| `talkrpp_install()` | Install the Python backend (conda). |
| `talkrpp_initialize()` | Activate an installed environment for this R session. |
| `find_talkrpp_env()` | Check whether a talk environment exists. |
| `list_talkrpp_envs()` | List the names of available Python environments. |
| `talkText()` | Transcribe audio to text. |
| `talkEmbed()` | Embed a whole audio file into one vector. |
| `talkTranscribeDiarise()` | Transcribe and split a recording by speaker. |
| `talkEmbedSegments()` | Embed each diarised segment (Whisper or WhiSPA). |


### Articles related to the talk-package
[WhiSPA: Semantically and Psychologically Aligned Whisper with Self-Supervised Contrastive and Student-Teacher Learning]() *Rajath Rao, Adithya V Ganesan, Oscar Kjell, Jonah Luby, Akshay Raghavan, Scott M. Feltman, Whitney Ringwald, Ryan L. Boyd, Benjamin J. Luft, Camilo J. Ruggero, Neville Ryant, ROMAN KOTOV, H. Schwartz*





